---
title: "stock price data AMSA"
author: "Ayrton Vitor"
date: "2022-10-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
spd <- read_csv("~/Downloads/T0804.csv", col_names = TRUE)
```
```{r}
head(spd)
summary(spd)
colSums(is.na(spd))
```
```{r}
ggplot(reshape2::melt(spd), mapping = aes(x = variable, y = value)) +
  geom_boxplot()
```

Same response variable, same scale, simmilar range, maybe we could proceed without
scaling.


### Sample cov matrix

```{r}
spd <- scale(spd)
d <- as.matrix(spd - colMeans(spd))
S <- (t(d) %*% d) / (nrow(spd) - 1)
print(S)
pc <- eigen(S)
print(pc)
```
```{r}
ggplot(data.frame(values=pc$values), aes(y = values, x = 1:5)) +
  geom_line() +
  labs(y="eigenvalue", x="principal component")
```
```{r}
sum(pc$values[1:2] / sum(pc$values))
sum(pc$values[1:3] / sum(pc$values))
```
The elbow suggest using only two principal components, but accounting for only
76\% of total variance, adding a third component to increase the total variance
could be important for subsequent pourpuses. We can take a look at the linear
correlations of principal components vs original variables.

```{r}
rho <- matrix(numeric(25), ncol = 5)
for (i in 1:5) {
  rho[i,] <- sqrt(pc$values[i]) * pc$vectors[,i] / sqrt(diag(S))
}
rho
```
With rows as principal components and columns as original values, we can notice
that Shell and Exxon Mobil are highly correlated with the first component,
whereas JP Morgan and Citibank, followed by Wells Fargo, appear to be the most
important variables to form the second component. No company seems particularly
highly correlated with the third principal component. This suggest that the first
component could be thought of as representing the oil companies, while the second
component, the banks.

Given that the gain in interpretability is not insanely high, as the contribution
of the banks to the first component and the contribution of oil companies to the
second component (negatively), is not exaclty negligible, we proceed with 3 variables
as it accounts for higher variance.

### Using princomp

```{r}
pca <- princomp(spd)
summary(pca)
pca$loadings
```
Notice that the variances are different.